Claret: Using Data Types for Highly Concurrent Distributed Transactions

### 2015-02-17 rmd/paper.Rmd
cite ellen selfie
tighten up some wording

abstract: |
    Out of the many NoSQL databases in use today, some that provide simple data structures for records, such as Redis and MongoDB, are now becoming popular. Building applications using these data structures rather than plain string values provides programmers with a way to communicate intent to the database system without sacrificing flexibility or committing to a fixed schema. Currently, this expressiveness is used to ensure related values are co-located so they can be quickly accessed together (e.g. maps and other aggregates) and to provide complex atomic operations (e.g. set insertion). However, there are many more ways in which data types can be used to make databases more efficient and simpler to use that are not yet being exploited.

    In this work, we demonstrate several ways of leveraging data structure semantics in databases, focusing primarily on commutativity. Reasoning about operation reordering can allow transactions to execute concurrently that would conflict under traditional concurrency control. Using Retwis, a Twitter clone built for Redis, as a case study, we show that using commutativity can reduce transaction abort rates for high-contention, update-heavy workloads that arise in real social networks. We conclude that data types are a good abstraction for database records, providing a safe and expressive programming model with ample opportunities for optimization, which will make databases more safe and scalable.
---
```{r setup, include=FALSE}
opts_chunk$set(dev='pdf', echo=F, message=F, warning=F, error=F, fig.width=3.2, fig.height=3)
```

@@

The move to non-relational (NoSQL) databases was motivated by a desire for scalability and flexibility. People found that by not enforcing strong consistency, they could better scale services to millions or billions of users while meeting tight performance goals. By forgoing the structure of relational schemas, they gained more direct control over performance at the cost of significant automatic optimization.
The observation has been that relaxed or eventual consistency is a good fit for many applications built on top of these NoSQL because potential conflicts are difficult to observe in practice and that users are likely to accept minor inconsistencies, such as two tweets being out of temporal order on their Twitter timeline. However, this leaves much to chance; there is likely no guarantee that more significant inconsistencies are impossible.

In situations where consistency really is crucial, developers can enforce stronger guarantees on top of any platform, but they often get it wrong. Recognizing the benefit to correctness and productivity, some systems have reintroduced serializable transactions, such as Google's Spanner \cite{Spanner}. However, this leaves programmers with two extreme choices with significant performance impact. If certain parts of an application can tolerate imprecision, then why not capture those properties in the database programming model? What if the programmer could express the semantics they desire succinctly and precisely, in a way that the database can better reason about conflicts and optimize performance, without giving up the flexibility and scalability of their existing systems?

Abstract data types (ADTs) are the solution. Rather than limiting the records in databases to being primitive types like strings or integers, raising them to more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a precise mechanism to express the intentions of programmers. Understanding the properties of ADT operations, such as which operations commute, can be used in a variety of ways to improve database performance, such as reducing the number of aborts. By choosing the right data type, programmers make tradeoffs between performance and precision based on their application's needs.

@@

Commutativity is well known, especially in distributed systems, for enabling important optimizations. Since the 80s, commutativity has been exploited by database systems designers \cite{Weihl:1988,Fekete:90}, within the safe confines of relational models, where knowledge of query plans and complete control of the data structures allows systems to determine when transactions may conflict. Recently, commutativity has seen a resurgence in systems without a predefined data model, such as NoSQL databases and transactional memory.
Commutativity is leveraged in eventually consistent databases for convergence in work such as RedBlue consistency \cite{Li:OSDI12} and conflict-free replicated data types (CRDTs) \cite{Shapiro:SSS11}. Other systems specialize for commutative operations to improve transaction processing, such as Lynx \cite{Zhang:SOSP13} for tracking serializability, and Doppel \cite{Narula:OSDI14} for executing operations in parallel on highly contended records, and HyFlow \cite{Kim:EuroPar} for reordering operations in the context of distributed transactional memory. We propose unifying and generalizing these uses of commutativity under the abstraction afforded by ADTs.

@@

**Conflict-free replicated data types (CRDTs)**, which were invented for eventual consistency, can actually be fit into our model as well, as a new kind of data type. This could allow copies to exist in different shards, asynchronously updating each other. By defining the same kind of *merge* function as traditional CRDTs, these replicas could ensure they all converge to the same state. Clients of these might find them more difficult to reason about, they might making this tradeoff in parts of the application where it otherwise cannot scale.

@@

\autoref{tput} shows the results of this simulation. When most of the traffic is content consumption (reading timelines), both systems perform well enough. However, when we simulate a workload where clients repost popular posts from their timelines, we see a viral propagation effect, where a large fraction of the users get and share a post. As Twitter came to a standstill when Ellen Degeneres's Oscar selfie set a retweeting record, so too does our baseline fall over. But with commutativity, performance continues to scale even under this highly contentious load.

>>>>

abstract: |
    Out of the many NoSQL databases in use today, some that provide simple data structures for records, such as Redis and MongoDB, are now becoming popular. Building applications out of these complex data types provides a way to communicate intent to the database system without sacrificing flexibility or committing to a fixed schema. Currently this expressiveness is used in limited ways, such as providing atomic operations or ensuring related values are co-located. However, there are many ways in which data types can be used to make databases more efficient and simpler to use that are not yet being exploited.

    In this work, we demonstrate several ways of leveraging abstract data type (ADT) semantics in databases, focusing primarily on commutativity. Using Retwis, a Twitter clone built for Redis, as a case study, we show that using commutativity can reduce transaction abort rates for high-contention, update-heavy workloads that arise in real social networks. We conclude that ADTs are a good abstraction for database records, providing a safe and expressive programming model with ample opportunities for optimization, which will make databases more safe and scalable.
---
```{r setup, include=FALSE}
opts_chunk$set(dev='pdf', echo=F, message=F, warning=F, error=F, fig.width=3.2, fig.height=3)
```

@@

The move to non-relational (NoSQL) databases was motivated by a desire for scalability and flexibility. People found that by giving up strong consistency, they could better scale services to millions or billions of users while meeting tight performance goals.
In fact for many applications, relaxed consistency feels natural; because of inherent uncertainty in timing and connectivity, users are likely to accept minor inconsistencies such as two tweets being out of temporal order or needing to retry an action.
However, this leaves much to chance: there is likely no guarantee that more significant inconsistencies are impossible.

When consistency truly is critical, developers can enforce stronger guarantees manually, or use serializable transactions in systems like Google's Spanner [@Spanner], but this leaves them with two extreme choices with a significant performance gap.
If certain parts of an application can tolerate imprecision, why not capture those properties in the programming model? Is there a way programmers can express the semantics they desire succinctly and precisely, helping the database optimize performance and scalability, without sacrificing flexibility?

We propose abstract data types (ADTs) as the solution. Rather than limiting the records in databases to primitive types like strings or integers, raising them to more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a precise mechanism to express the intentions of programmers. We demonstrate one way of using commutativity of ADT operations to reduce transaction aborts and explore several other ways of leveraging commutativity and data types to improve database performance and allow programmers to make tradeoffs between performance and precision.

@@

Commutativity is well known, especially in distributed systems, for enabling important optimizations. Since the 80s, commutativity has been exploited by database systems designers \cite{Weihl:1988,Fekete:90}, within the safe confines of relational models, where knowledge of query plans and complete control of the data structures allows systems to determine when transactions may conflict. Recently, commutativity has seen a resurgence in systems without a predefined data model, such as NoSQL databases and transactional memory.
Commutativity is leveraged in eventually consistent databases for convergence in work such as RedBlue consistency \cite{Li:OSDI12} and conflict-free replicated data types (CRDTs) \cite{Shapiro:SSS11}. Other systems specialize for commutative operations to improve transaction processing, such as Lynx \cite{Zhang:SOSP13} for tracking serializability, and Doppel \cite{Narula:OSDI14} for executing operations in parallel on highly contended records, and HyFlow \cite{Kim:EuroPar13} for reordering operations in the context of distributed transactional memory. We propose unifying and generalizing these uses of commutativity under the abstraction afforded by ADTs.

@@

**Conflict-free replicated data types (CRDTs)**, which were invented for eventual consistency, can actually be fit into our model as well, as a new kind of data type. This could allow copies to exist in different shards, asynchronously updating each other. By defining the same kind of *merge* function as traditional CRDTs, these replicas could ensure they all converge to the same state. Clients of these might find them more difficult to reason about but may consider making this tradeoff in parts of the application where it otherwise cannot scale.

@@

\autoref{tput} shows the results of this simulation. When most of the traffic is content consumption (reading timelines), both systems perform well enough. However, when we simulate a workload where clients repost popular posts from their timelines, we see a viral propagation effect, where a large fraction of the users get and share a post. As Twitter came to a standstill when Ellen DeGeneres's Oscar selfie set a retweeting record \cite{ellenselfie}, so too does our baseline fall over. But with commutativity, performance continues to scale even under this highly contentious load.
